Ludwig User Tutorial

### 2018-10-22 docs/tutorial/tutorial.tex
Amended tutorial to reflect latest changes in post-processing

\usepackage{amsmath}
\usepackage{moreverb}
\usepackage{lscape}
\usepackage{epic}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{float}
\graphicspath{{./tutorialFigs/}}

@@

The system requirements are an up-to-date version of a C compiler as well as working installations 
of a text editor like \texttt{vi} or \texttt{emacs}, the visualisation tool
ParaView \\
(\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.

@@

The \texttt{Ludwig} project is currently hosted at our central repository at GitHub:\\
(\hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}).
\smallskip

@@

A full documentation of \texttt{Ludwig} can be obtained by running 
the command \texttt{make pdf} twice in the \texttt{/docs} directory:

@@

@@

OPTS = -DNP_D3Q6 -DNDEBUG
LIBS = -L../target -ltarget -lm
INC = -I. -I ../target
...
\end{lstlisting}
\item To compile the code in serial issue: \\
\begin{lstlisting}
$ make serial
\end{lstlisting} 
To compile the parallel version of the code issue: \\
\begin{lstlisting}
$ make mpi
\end{lstlisting} 
This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
Before going from a serial to a parallel compilation or vice versa
you should clean the directory by issuing
\begin{lstlisting}
$ make clean
\end{lstlisting}

@@

In this section the Poiseuille flow of a simple fluid in a cavity, is outlined. 
The example is a step-by-step guide for performing a short run ($10,000$ time-steps) 
with \texttt{Ludwig} code, followed by a short introduction to the post-processing 
and data analysis procedures. Before starting, the code should be compiled in 
parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory \texttt{/trunk/docs/tutorial/test1} 
to your working directory. Note that this input file is a modified version of the 
reference input file \texttt{input.ref} located in \texttt{/trunk/src} with a limited set of parameters
for this run. 
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 4 ./Ludwig.exe input
\end{lstlisting}
\item You should now see statistics being reported on your standard output and 
output files being created with names of the form \texttt{vel-XXXXXXXX.001-001}. 
These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
\item The first step to the post-processing of these data is to copy the files \texttt{vtk\_extract.c} 
and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util} into your working directory.
\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
\begin{enumerate}
\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file),
\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format,
\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k}, 
\item \texttt{vtk\_header=0} \\ to omit headers (these will be required later for visualisation in \textit{ParaView}),
\item \texttt{output\_cmf=0} to produce data in row-major format (column-major format is required for \textit{ParaView}).
\end{enumerate}
\item Edit \texttt{vtk\_extract\_script.py}. Set:
\begin{enumerate}
\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and increments of the velocity data
to be processed.
\item \texttt{vel=1} for velocity post-processing, 
\item The other flags should be set to zero, e.g. \texttt{phi=0}. 
\end{enumerate}
\item Issue the command:
\begin{lstlisting}
$ python vtk_extract_script.py
\end{lstlisting}
\item You should now see files named \texttt{vel-XXXXXXXX.vtk} (Note: The extension \texttt{.vtk} is only used here for simplicity.
It does not mean the generated data is in vtk-format). 
The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\

To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
the spatial $x$-coordinate on the lattice with the command:
\begin{lstlisting}
gnuplot> p`vel-XXXXXXXX.vtk' u 1:5
\end{lstlisting} 
By plotting the data in each file, you should be able to see the velocity profile converging to the final profile 
as shown in Fig. \ref{fig:velocityProfile}.

In the directory \texttt{/trunk/docs/tutorial/test1} a reference file of the final 
output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
\end{enumerate}

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/trunk/docs/tutorial/test2} 
to your working directory. This input file contains the parameters required for this run. 
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item As before, you should now see statistics being reported on your standard output but 
in this case two sets of output files being created containing the velocity and compositional
order parameter ($\phi$) data at each lattice point in the simulation. In this example, these 
data are in binary format.
\item The post-processing of these data again requires the files \texttt{vtk\_extract.c} 
and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util}.
\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
\begin{enumerate}
\item \texttt{input\_binary\_=1} \\ if the velocity and $\phi$ data are in binary format (see 
flag in input file),
\item \texttt{output\_binary\_=0} \\ as the output of the post-processing routines should be 
in ASCII format,
\item \texttt{output\_index=0} \\ as Cartesian lattice indices are not required, 
\item \texttt{vtk\_header=1} \\ to include headers,
\item \texttt{output\_cmf=1} to produce data in column-major format.
\end{enumerate}
\item Edit \texttt{vtk\_extract\_script.py}. Set:
\begin{enumerate}
\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and 
increments of the velocity data
to be processed.
\item \texttt{vel=1} for velocity post-processing, 
\item \texttt{phi=1} for $\phi$ post-processing,
\item The other flags should be set to zero. 
\end{enumerate}
\item Issue the command:
\begin{lstlisting}
$ python vtk_extract_script.py
\end{lstlisting}
\item You should now see files with the extension \texttt{.vtk} corresponding to the velocity 
and $\phi$ data ready to be visualised using ParaView.
\begin{enumerate}
\item Open the files containing the $\phi$ data.
\item Click on the group of files in the pipeline browser and apply a contour map using 
Filters/Common/Contour or in the tool bar:

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/trunk/docs/tutorial/test3} to your working directory. We see in this input file that 
there are many more parameters 
than in the previous examples. Specifically, we note that \texttt{colloid\_cell\_min} 
must be set if colloids are present and must be at least $2r + \delta$ where $r$ is colloid 
radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.

@@

\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly compile the code in serial (see section \ref{configbuild})and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the file required for the initialisation and visualisation of colloids respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file, \texttt{config.cds.init.001-001} which should be copied into your working directory.
\item In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
\item \texttt{cds\_with\_v  = 1} \\ which will output the velocities and positions of the colloids,
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory along with the 
usual post-processing scripts, \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} 
scripts. 
\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item For the post-processing of this system, in \texttt{vtk\_extract.c}, set:
\begin{enumerate}
\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
\item \texttt{output\_lc\_op\_   = 1} \\ to visualise the liquid crystal order parameter,
\item \texttt{output\_lc\_dir\_  = 1} \\ to visualise the liquid crystal director field,
\item \texttt{output\_lc\_biop\_ = 0} \\ since the liquid crystals are not biaxial,
\end{enumerate}
\item and in \texttt{vtk\_extract\_script.py},
\begin{enumerate}
\item \texttt{vel = 1}, \\ to visualise the velocity field,
\item \texttt{q = 1}, \\ since there are liquid crystals present,
\item \texttt{colcds = 0} and 
\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
\end{enumerate}
\item After running the python script as before, \texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{op-q-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:

@@

\item For the director field open the \texttt{dir-q-XXXXXXXX.vtk} and apply the `glyphvector' filter. In this case, change the glyph 
type from `arrow' to `line' and scale the colour of the lines by the $x$-component of the 
vector.
\item The velocity field can be visualised using arrows as before with the `glyphvector' filter.
\item To visualise the colloids, open the files \texttt{col-cdsXXXXXXXX.csv}. This will open 
a table, which can be closed. Next apply the filter `Table to Points' found in 
Filters/Alphabetical/Table to Points and set the X, Y and Z Columns in the `Properties' tab 
below the Pipeline Browser to be the $x$, $y$ and $z$-coordinates in the table via the 
drop down menus:

>>>>

\usepackage{amsmath}
\usepackage{moreverb}
\usepackage{lscape}
\usepackage{epic}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{float}
\usepackage{hyperref}
\graphicspath{{./tutorialFigs/}}

@@

The system requirements are an up-to-date version of a C compiler as well as working installations 
of a text editor like \texttt{vi} or \texttt{emacs}, the visualisation tool
ParaView \\
({\color{blue}\hyperref[ParaView]{https://www.paraview.org}}), the MPI-library 
for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.
The instructions for compilation of this tutorial and the full documentation assume a LaTeX distribution
and \texttt{pdflatex} are installed.

@@

The \texttt{Ludwig} project is currently hosted at our central repository at GitHub:\\
({\color{blue} \hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}}).
\smallskip

@@

A full documentation of \texttt{Ludwig} can be obtained by running 
the command \texttt{make pdf} twice (for references) in the \texttt{/docs} directory:

@@

This tutorial and directories with input files and sample output is contained in 
a subdirectory of \texttt{/docs}. It can be compiled in the following way:

\begin{lstlisting}[style=terminalverbatim]
$ cd /ludwig/docs/tutorial
$ make
\end{lstlisting}

@@

OPTS = -DNP_D3Q6 -DNDEBUG
LIBS = -L../target -ltarget -lm
INC = -I. -I ../target
...
\end{lstlisting}
\item To compile the code in serial issue: \\
\begin{lstlisting}
$ make serial
\end{lstlisting} 
To compile the parallel version of the code issue: \\
\begin{lstlisting}
$ make mpi
\end{lstlisting} 
This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
Note that more than one thread can be used in the compilation process,
which can lead to a considerable speedup  
\begin{lstlisting}
$ make -j4 mpi
\end{lstlisting} 
Before going from a serial to a parallel compilation or vice versa
you should clean the directory by issuing
\begin{lstlisting}
$ make clean
\end{lstlisting}

@@

In this section the Poiseuille flow of a simple fluid in a cavity, is outlined. 
The example is a step-by-step guide for performing a short run ($10,000$ time-steps) 
with \texttt{Ludwig} code, followed by a short introduction to the post-processing 
and data analysis procedures. Before starting, the code should be compiled in 
parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.
\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory \texttt{/docs/tutorial/test1} 
to your working directory. Note that the input files in this tutorial are modified versions of the 
reference input file \texttt{input.ref} in \texttt{/src} with a limited set of parameters
for each run.
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 4 ./Ludwig.exe input
\end{lstlisting}

\item You should now see statistics being reported on your standard output and 
output files being created with names of the form \texttt{vel-XXXXXXXX.001-001}. 
These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
\item The first step to the post-processing of the data is to compile the \texttt{extract.c} 
routine in \texttt{/util}. 
\item Edit \texttt{extract.c}. Specifically, set the flags for:
\begin{enumerate}
\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file for \texttt{test1}),
\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format for this tutorial,
\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k} for direct visualisation with \texttt{gnuplot}, 
\end{enumerate}
\item Compile with 
\begin{lstlisting}
$ make extract 
\end{lstlisting} 
and copy the executable \texttt{extract} into your working directory.
\item Post-process the raw data by issuing 
\begin{lstlisting}
$ ./extract meta_file_name data_file_name_stub
\end{lstlisting} 
replacing the filenames with those of the present case (i.e. \texttt{vel.001-001.meta} and e.g. \texttt{vel-00010000}).
\item You should now see files named \texttt{vel-XXXXXXXX}.
The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\
\item To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
the spatial $x$-coordinate on the lattice with the command:
\begin{lstlisting}
gnuplot> p `vel-XXXXXXXX' u 1:5
\end{lstlisting} 
By plotting the data in each file, you should be able to see the velocity profile converging to the final profile as shown in Fig. \ref{fig:velocityProfile}.

In the directory \texttt{/docs/tutorial/test1} a reference file of the final 
output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
\end{enumerate}

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test2} 
to your working directory. This input file contains the parameters required for this run. 
\item Run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item As before, you should now see statistics being reported on your standard output. But 
in this case two sets of output files are being created, one containing the velocity and 
the other compositional order parameter ($\phi$) data at each lattice point in the simulation. 
In this example, these data are in binary format.
\item The post-processing of the data requires again a compiled exectuable of \texttt{extract.c} in
\texttt{/util}, but this time with slightly different settings. Specifically, set the flags for:
\begin{enumerate}
\item \texttt{input\_binary\_=1} \\ if the scalar order parameter and velocity data are in binary format (see 
flag in input file for \texttt{test2}),
\item \texttt{output\_binary\_=0} \\ as the output of the post-processing routines should be 
in ASCII format,
\item \texttt{output\_index=0} \\ as Cartesian lattice indices are not required, 
\item \texttt{output\_cmf=1} to produce data in column-major format, which is required for the correct 
visualisation with ParaView.
\end{enumerate}
\item Compile with 
\begin{lstlisting}
make extract
\end{lstlisting} 
and copy the executable \texttt{extract} into your working directory.
\item A python script \texttt{extract.py}
can be used for convenient post-processing of multiple raw data files and time steps.
 Edit \texttt{extract.py} and set:
\begin{enumerate}
\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and 
increments of the velocity data
to be processed.
\item \texttt{vel=1} for velocity post-processing, 
\item \texttt{phi=1} for $\phi$ post-processing,
\item The other flags should be set to zero. 
\end{enumerate}
\item Issue the command:
\begin{lstlisting}
$ python extract.py
\end{lstlisting}
\item You should now see files with the extension \texttt{.vtk} corresponding to the velocity 
and $\phi$ data ready to be visualised using ParaView.
\begin{enumerate}
\item In Paraview open the files containing the $\phi$ data.
\item Click on the group of files in the pipeline browser and apply a contour map using 
Filters/Common/Contour or in the tool bar:

@@

\begin{enumerate}
\item Copy the input file \texttt{input} located in the directory 
\texttt{/docs/tutorial/test3} to your working directory. We see in this input file that 
there are many more parameters 
than in the previous examples. Specifically, we note that \texttt{colloid\_cell\_min} 
must be set if colloids are present and must be at least $2r + \delta$ where $r$ is colloid 
radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.

@@

\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}. There you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the files required for the initialisation and post-processing of colloids, respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.
\item In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
\item \texttt{cds\_with\_v  = 1} \\ which will output the positions and velocity data of the colloids,
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory.
\item For the post-processing of this system, in \texttt{extract.c}, set:
\begin{enumerate}
\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
\end{enumerate}
compile in the usual manner and copy the post-processing executable \texttt{extract} and 
the script \texttt{extract.py} as well into your working directory. 
\item In \texttt{extract.py},
\begin{enumerate}
\item \texttt{vel = 1}, \\ to visualise the velocity field,
\item \texttt{q = 1}, \\ since there are liquid crystals present,
\item \texttt{colcds = 0} and 
\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
\end{enumerate}

\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}

\item Run the post-processing routine as before using the python script. 
\texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{lcs-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:

@@

\item For the director field open the \texttt{lcd-XXXXXXXX.vtk} and apply the `glyphvector' filter. In this case, change the glyph 
type from `arrow' to `line' and scale the colour of the lines by the $x$-component of the 
vector.
\item The velocity field can be visualised using arrows as before with the `glyphvector' filter.
\item To visualise the colloids, open the files \texttt{col-cdsXXXXXXXX.csv}. This will open 
a table, which can be closed. Next apply the filter `Table to Points' found in 
Filters/Alphabetical/Table to Points and set the X, Y and Z Columns in the `Properties' tab 
below the Pipeline Browser to be the $x$, $y$ and $z$-coordinates in the table via the 
drop down menus:
